---
title: "Solution for Diagnostic Quiz"
author: "Teaching Team"
date: "13/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 5

_Write programs or script using the programming language of your choice that perform the following tasks.  Use programming primitives as opposed to dedicated library functions:_

- (a) Given an input matrix, check if it is:  Symmetric, upper-triangular, lower-triangular.

Symmetric
```{r}
isSymmetric(D3 <- diag(3))
D3
isSymmetric(D4 <- matrix(rnorm(9),nrow=3,ncol=3))
D4
```


upper-triangular and  lower-triangular.
```{r}
library(Matrix)
D4[upper.tri(D4)] <- 0
D4
isTriangular(D4,upper=FALSE)
isTriangular(D4,upper=TRUE)
```



- (b)  Given a list of numbers, return a sorted list of numbers


```{r}
x <- rnorm(10)*10
x
cat("x sorted \n")
sort(x)
```

- (c)  Given  a  sequence  ofninput  vectors,  each  of  length $p$,  compute  the $p\times p$ sample covariance matrix of these vectors.

```{r}
p <- 8
set.seed(10)
X <- data.frame(X1=rnorm(p),X2=runif(p),X3=rt(p,df=1)*10)
X
cat("Covariance matrix  of X1, X2, X3\n")
cov(X)
cat("Correlation matrix  of X1, X2, X3\n")
cor(X)
```

Primitive function

```{r}
cat("Covariance matrix  of X1, X2, X3\n")
X.center <- scale(X,center=TRUE,scale=FALSE)
(1/(p-1))*t(X.center)%*%X.center
cat("Correlation matrix  of X1, X2, X3\n")
X.c.s <- scale(X,center=TRUE,scale=TRUE)
(1/(p-1))*(t(X.c.s)%*%X.c.s)
```



- (d)  Use brute-force Riemann sums to illustrate numerically that

\[
\int_{x_1 = -5}^5 \int_{x_2= -5}^5 ~\frac{1}{2 \pi}e^{-\frac{x_1^2 + x_2^2}{2}} ~ dx_1\, dx_2  \approx 1.
\]

```{r}
library(cubature)
integrand <- function(x) (1/(2*pi))*exp(-0.5*(x[1]^2+x[2]^2))
adaptIntegrate(integrand, lowerLimit = c(-5,-5), upperLimit = c(5, 5))
```

```{r}
x1lower <- x2lower <- -5
x1upper <- x2upper <- 5
n <- 1000
deltak <- (x2upper - x2lower)/n
Deltak <- deltak**2
center.x <- c(0,0)
f.center <- 0
for (i in 1:n) {
  for (j in 1:n){
    center.x[1] <- x1lower + i*deltak - deltak/2
    center.x[2] <- x2lower + j*deltak - deltak/2
  f.center <- f.center+ integrand(center.x)
  }
}
integral <- Deltak*f.center
integral
```

