---
title: Neural Network For Classification
author: 
- name:  Benoit and Sarat and Yoni 
  affiliation: Macquarie University
  email: benoit.liquet-weiland@mq.edu.au
date: '`r doc_date()`'
output:
  BiocStyle::html_document:
    toc_depth: 2
  BiocStyle::pdf_document:
    toc: true
    keep_tex: yes
    includes:
     in_header: peamble.tex
    toc_depth: 2
fontsize: 15pt
header-includes:
- \usepackage{color}
---



[//]: A few LaTeX Macros
$\newcommand{\vect}[1]{\boldsymbol{#1}}$
$\newcommand{\transp}{^{\text{T}}}$
$\newcommand{\mat}[1]{\boldsymbol{\mathcal{#1}}}$
$\newcommand{\sign}{\text{sign}}$


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = NA,comment = NA)
library(knitr)
```



# Logistic Regression by Scratch: Binary example  (HW1)

## Data Set: A Flower

First, let’s get the dataset you will work on. The following code will load a "flower" 2-class dataset.


```{r,message=FALSE}
library(ggplot2)
library(dplyr)
load("planar_solar.RData")
df <- planarsolar
df <- df[sample(nrow(df)), ]
head(df)
df %>%
  ggplot(aes(x = x1, y= x2,colour=factor(y))) + geom_point()
```



### Train-Test Split

- Split it into train and test sets.

- 80\% of our data into our train set and the remaining 20\% into our test set.


```{r}
set.seed(69)
train_test_split_index <- 0.8 * nrow(df)

train <- df[1:train_test_split_index,]
head(train)

test <- df[(train_test_split_index+1): nrow(df),]
head(test)
```




### Preprocess

- Neural networks work best when the input values are standardized. 

- Scaled all the values to to have their mean = 0 and standard-deviation =1.

- Standardizing input values speeds up the training and ensures faster convergence.

However, in this tutorial we decided to not scaled the values to compared to the classical **GLM**.

```{r}
#X_train <- scale(train[, c(1:2)])
X_train <- train[, c(1:2)]

y_train <- train$y
dim(y_train) <- c(length(y_train), 1) # add extra dimension to vector

#X_test <- scale(test[, c(1:2)])
X_test <- test[,c(1:2)]
y_test <- test$y
dim(y_test) <- c(length(y_test), 1) # add extra dimension to vector
```





Now we convert our input and output to matrices from data frames. Converting to a matrix format often speed up the computations.

```{r}
X_train <- as.matrix(X_train, byrow=TRUE)
X_train <- t(X_train)
dim(X_train)

y_train <- as.matrix(y_train, byrow=TRUE)
y_train <- t(y_train)
dim(y_train)

X_test <- as.matrix(X_test, byrow=TRUE)
X_test <- t(X_test)
dim(X_test)
y_test <- as.matrix(y_test, byrow=TRUE)
y_test <- t(y_test)
dim(y_train)
```

## Logistic Regression with a Neural Network


The main steps for building a Neural Network are:

1. Define the model structure (such as number of input features) 

2. Initialize the model's parameters

3. Loop:
    - Calculate current loss (forward propagation)
    - Calculate current gradient (backward propagation)
    - Update parameters (gradient descent)

### Sigmoid function 

```{r}
sigmoid <- function(x){
  return(1 / (1 + exp(-x)))
}
```


###  Function Layer Size

We first build a function which provides the size (number of neurons) of all the layers in our neural-net. This function will be more usefull for Neural Network with hidden layer.



```{r}
getLayerSize <- function(X, y) {
  n_x <- dim(X)[1]  # number of neuron in input layer
  n_y <- dim(y)[1]  # number of neuron output layer
  
  size <- list("n_x" = n_x,
               "n_y" = n_y)
  
  return(size)
}

```

### Parameter Initialisation 

The function _initializeParameters()_ takes as argument an input matrix and a list which contains the layer sizes i.e. number of neurons. The function returns the trainable parameters $W1$, $b1$, $W2$, $b2$.


```{r}
initializeParameters <- function(X, list_layer_size){
  
  n_x <- list_layer_size$n_x
  n_y <- list_layer_size$n_y
  
  W1 <- matrix(runif(n_y * n_x), nrow = n_y, ncol = n_x, byrow = TRUE) * 0.01
  b1 <- matrix(rep(0, n_y), nrow = n_y)
 
  params <- list("W1" = W1,
                 "b1" = b1)
  return (params)
}

```

### Forward Propagation Step 

```{r}
forwardPropagation <- function(X, params, list_layer_size){
  
  m <- dim(X)[2]
  n_y <- list_layer_size$n_y
  
  W1 <- params$W1
  b1 <- params$b1
 
  b1_new <- matrix(rep(b1, m), nrow = n_y)
 
  Z1 <- W1 %*% X + b1_new
  A1 <- sigmoid(Z1)

  cache <- list("Z1" = Z1,
                "A1" = A1)
  return (cache)
}
```

### Compute cost

```{r}
computeCost <- function(X, y, cache) {
  m <- dim(X)[2]
  A1 <- cache$A1
  logprobs <- (log(A1) * y) + (log(1-A1) * (1-y))
  cost <- -sum(logprobs/m)
  return (cost)
}
```

### Backpropagation

```{r}
backwardPropagation <- function(X, y, cache, params, list_layer_size){
  m <- dim(X)[2]
  n_x <- list_layer_size$n_x
  n_y <- list_layer_size$n_y
  
  A1 <- cache$A1
  W1 <- params$W1
  
  dZ1 <- A1 - y
  dW1 <- 1/m * (dZ1 %*% t(X)) 
  db1 <- matrix(1/m * sum(dZ1), nrow = n_y)
  db1_new <- matrix(rep(db1, m), nrow = n_y)
  
  grads <- list("dW1" = dW1, 
                "db1" = db1)
  
  return(grads)
}

```

### Update Parameters


```{r}
updateParameters <- function(grads, params, learning_rate){
  
  W1 <- params$W1
  b1 <- params$b1
  
  dW1 <- grads$dW1
  db1 <- grads$db1
  
  W1 <- W1 - learning_rate * dW1
  b1 <- b1 - learning_rate * db1
  
  updated_params <- list("W1" = W1,
                         "b1" = b1)
  return (updated_params)
}
```


## Train our model

```{r}
trainModel <- function(X, y, num_iteration, lr){
  
  layer_size <- getLayerSize(X, y)
  init_params <- initializeParameters(X, layer_size)
  
  cost_history <- c()
  
  for (i in 1:num_iteration) {
    fwd_prop <- forwardPropagation(X, init_params, layer_size)
    cost <- computeCost(X, y, fwd_prop)
    back_prop <- backwardPropagation(X, y, fwd_prop, init_params, layer_size)
    update_params <- updateParameters(back_prop, init_params, learning_rate = lr)
    init_params <- update_params
    
    cost_history <- c(cost_history, cost)
    
    if (i %% 10 == 0){ cat("Iteration", i, " | Cost: ", cost, "\n")
      cat("Iteration", i, " | W1: ", update_params$W1, "\n")
    }
  }
  
  model_out <- list("updated_params" = update_params,
                    "cost_hist" = cost_history)
  
  return (model_out)
}
```

## Application on the data set 


### Logistic Neural Network



```{r,eval=TRUE}
layer_size <- getLayerSize(X_train, y_train)
layer_size


init_params <- initializeParameters(X_train, layer_size)
lapply(init_params, function(x) dim(x))
fwd_prop <- forwardPropagation(X_train, init_params, layer_size)

lapply(fwd_prop, function(x) dim(x))

cost <- computeCost(X_train, y_train, fwd_prop)
cost

back_prop <- backwardPropagation(X_train, y_train, fwd_prop, init_params, layer_size)
lapply(back_prop, function(x) dim(x))


update_params <- updateParameters(back_prop, init_params, learning_rate = 0.01)
lapply(update_params, function(x) dim(x))


EPOCHS = 100
LEARNING_RATE = 0.9

train_model <- trainModel(X_train, y_train, num_iteration = EPOCHS, lr = LEARNING_RATE)
```

The parameters are 

```{r}
train_model$updated_params
```

## Logistic Model using GLM

```{r}
lr_model <- glm(y ~ x1 + x2, data = train,family=binomial())
lr_model
```
## Performance of our binary classifier 


- Prediction function for new data

```{r}
makePrediction <- function(X, y){
  layer_size <- getLayerSize(X, y)
  params <- train_model$updated_params
  fwd_prop <- forwardPropagation(X, params, layer_size)
  pred <- fwd_prop$A1
  return (pred)
}

```

### confusion matrix

```{r}
y_pred <- makePrediction(X_test, y_test)
y_pred <- round(y_pred)

tb_nn <- table(y_test, y_pred)
tb_nn
```


### Accuracy Metrics

```{r}
calculate_stats <- function(tb, model_name) {
  acc <- (tb[1] + tb[4])/(tb[1] + tb[2] + tb[3] + tb[4])
  recall <- tb[4]/(tb[4] + tb[3])
  precision <- tb[4]/(tb[4] + tb[2])
  f1 <- 2 * ((precision * recall) / (precision + recall))
  
  cat(model_name, ": \n")
  cat("\tAccuracy = ", acc*100, "%.")
  cat("\n\tPrecision = ", precision*100, "%.")
  cat("\n\tRecall = ", recall*100, "%.")
  cat("\n\tF1 Score = ", f1*100, "%.\n\n")
}
```


```{r}
calculate_stats(tb_nn,"Logistic Neural network")
```


## Decision Boundaries

We firstly make a grid of points around the input space:

```{r}
grid <- expand.grid(x1 = seq(min(df$x1) - 1,
                             max(df$x1) + 1,
                             by = .25),
                    x2 = seq(min(df$x2) - 1,
                             max(df$x2) + 1,
                             by = .25))
```


Then feed these points forward through our trained neural network.

```{r}
newdata <- t(data.matrix(grid[, c('x1', 'x2')]))

fwd_prop <- forwardPropagation(newdata, train_model$updated_params, layer_size)

grid$class <- factor((fwd_prop$A1 > .5) * 1,
                     labels = levels(factor(df$y)))

```


```{r,message=FALSE}
library(tidyverse)
class <- as.factor(df$y)
 df %>% add_column(class=class)
  ggplot(df)+ aes(x1,x2,colour=class)  +
  geom_point(data = grid, size = .5) +
  geom_point() +
  labs(x = expression(x[1]), y = expression(x[2]))
```



# Adding one Hidden Layer

## feedforward  

To write a compact code, we include the biais parameter in the weight matrices parameters. Just to play, we use the sigmoid activation function in the hidden layer. The feedforward function is very compact:

```{r}
feedforward <- function(x, w1, w2) {
  z1 <- cbind(1, x) %*% w1
  h <- sigmoid(z1)
  z2 <- cbind(1, h) %*% w2
  list(output = sigmoid(z2), h = h)
}
```

Using the computation graph, the backpropagation function could be easily written as

```{r}
backpropagate <- function(x, y, y_hat, w1, w2, h, learn_rate) {
  dw2 <- t(cbind(1, h)) %*% (y_hat - y)
  dh  <- (y_hat - y) %*% t(w2[-1, , drop = FALSE])
  dw1 <- t(cbind(1, x)) %*% (h * (1 - h) * dh)
  
  w1 <- w1 - learn_rate * dw1
  w2 <- w2 - learn_rate * dw2
  
  list(w1 = w1, w2 = w2)
}
```

We can now wrap it to get our train model function:

```{r}
train <- function(x, y, hidden = 5, learn_rate = 1e-2, iterations = 1e4) {
  d <- ncol(x) + 1
  w1 <- matrix(rnorm(d * hidden), d, hidden)
  w2 <- as.matrix(rnorm(hidden + 1))
  for (i in 1:iterations) {
    ff <- feedforward(x, w1, w2)
    bp <- backpropagate(x, y,
                        y_hat = ff$output,
                        w1, w2,
                        h = ff$h,
                        learn_rate = learn_rate)
    w1 <- bp$w1; w2 <- bp$w2
  }
  list(output = ff$output, w1 = w1, w2 = w2)
}
```

```{r,cache=TRUE,cache.path='cache/'}
#x <- data.matrix(df[, c('x1', 'x2')])
#y <- df$y
#y <- as.numeric(y_train)
nnet10 <- train(t(X_train), as.numeric(y_train), hidden = 10, iterations = 1e5)
```

### Is it better ?

```{r}
mean((nnet10$output > .5) == t(y_train))
```

### What about on the test set ?

```{r}
predout <- feedforward(x = t(X_test),w1 = nnet10$w1,w2 = nnet10$w2)
y_pred_nnet10<- (predout$output > .5) * 1
tb_nnH <- table(y_test, y_pred_nnet10)
tb_nnH
```

## confusion matrix

```{r}
calculate_stats(tb_nnH,"Neural network with one Hidden layer and 10 neurons")
```



## Visualisation

```{r}
ff_grid <- feedforward(x = data.matrix(grid[, c('x1', 'x2')]),
                       w1 = nnet10$w1,
                       w2 = nnet10$w2)
class <- as.factor(df$y)
grid$class <- factor((ff_grid$output > .5) * 1,
                     labels = levels(class))

```

```{r}
class <- as.factor(df$y)
 df %>% add_column(class=class)
  ggplot(df)+ aes(x1,x2,colour=class)  +
  geom_point(data = grid, size = .5) +
  geom_point() +
  labs(x = expression(x[1]), y = expression(x[2]))
```

# Neural Network using Software


## R packages for deep learning 

- **nnet**: Package for feedforward neural networks with a single hidden layer and multinomial log-linear models

- **neuranet**: Training of neural networks using backpropagation

- **tensorflow**: Interface to TensorFlow

- **deepnet**: Deep learning toolkit in R

- **rnn**: Package to implement Recurrent Neural Networks

more here https://cran.r-project.org/web/views/MachineLearning.html


## Keras

The R package **keras** is an interface to the Python **keras** package.

### Install Keras

```{r,eval=FALSE}
install.packages(keras)
```

### Load Keras

```{r}
library(keras)
```

### Defining a keras Model

Keep in mind in you are lost 

```{r,eval=FALSE}
help(package = keras)
```

Here, we use a fully connected neural network. 

- First we initialize a sequential model using **keras_model_sequential()** function

- Sequential model is a linear stack of layers. Use **keras_model_sequential()** function to add a series of layer functions.

- Specify activation functions: **relu**, **softmax**, **sigmoid**, **tanh**, ...

- Need to specify the loss function: **mean_squared_error**, **binary_crossentropy**, **categorical_crossentropy**

- Optimizer could be: **sgd**, **adam**, **rmsprop**, **Adadelta**

- learning rate specify with **lr**

### Binary classifier with one hidden layer

```{r,comment=FALSE}
set.seed(1)
model <- keras_model_sequential()
model %>%
layer_dense(units = 8, activation = "relu", input_shape = c(2)) %>%
#layer_dense(units = 10, activation = "sigmoid") %>%
layer_dense(units = 1,activation = "sigmoid") %>%
compile(
optimizer = "adam",
loss = "binary_crossentropy",
metrics = c("accuracy")
)
summary(model)
```


## Run the model on the full data

### Scale the data

```{r}
X_train <- scale(df[, c(1:2)])
#X_train <- train[, c(1:2)]
y_train <- df$y
#dim(y_train) <- c(length(y_train), 1) # add extra dimension to vector
#X_test <- scale(test[, c(1:2)])
#X_test <- test[,c(1:2)]
#y_test <- test$y
#dim(y_test) <- c(length(y_test), 1) # add extra dimension to vector
```




```{r,cache=TRUE,cache.path='cache/'}
history <- model %>% fit(X_train, y_train, epochs = 500, batch_size = 40,validation_split = 0)
model %>% evaluate(X_train, y_train)
```

```{r}
plot(history)
```

```{r}
keras_pred_train <- model %>% predict_classes(X_train)
table(Predicted = keras_pred_train, Actual = t(y_train))
```

## Need more iteration


```{r,eval=FALSE}
history <- model %>% fit(X_train, y_train, epochs = 10000, batch_size = 40,validation_split = 0)
plot(history)
model %>% evaluate(X_train, y_train)
```


### Overfitting ?

Reproduce similar analysis using 80$\%$ for train data and evaluate on  test set (20$\%$)

```{r,eval=FALSE}
keras_pred_test <- model %>% predict_classes(X_test)
table(Predicted = keras_pred_test, Actual = y_test)
```

## Explore different models

- Add more neurons
- Try the **tanh** activation function 
- Change the optimizer function: **sgd**

